---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wex

<!-- badges: start -->
[![](https://www.r-pkg.org/badges/version/wex?color=green)](https://cran.r-project.org/package=wex) [![](https://img.shields.io/github/last-commit/timginker/wex.svg)](https://github.com/timginker/wex/commits/master)
[![](http://cranlogs.r-pkg.org/badges/grand-total/wex)](https://cran.r-project.org/package=wex)
<!-- badges: end -->


`wex` is an R package for computing exact observation weights for the Kalman filter and smoother, following Koopman and Harvey (2003). The package provides tools for analyzing linear Gaussian state-space models by quantifying how individual observations contribute to filtered and smoothed state estimates.

These weights are particularly useful in applications such as dynamic factor models, where they can be used to decompose latent factors into contributions from observed variables (see Example 2 below).


## Installation

The stable version of `wex` can be installed from CRAN:

``` r
install.packages("wex")
```

The development version of `wex` can be installed from [GitHub](https://github.com/):

``` r
# install.packages("devtools")
devtools::install_github("timginker/wex")
```

## Example 1: Local level model

In this illustrative example, we fit a local level model to the `Nile` data and compute the corresponding filtered and smoothed state estimates. We then extract the observation weights and use them to reconstruct these estimates from the observed data.

The resulting estimates are shown in the plot below.

```{r echo=F, eval=T}
library(FKF)
library(wex)
library(kableExtra)
data(Nile)
y <- Nile
y[c(3, 10)] <- NA # NA values can be handled

# Computing smoothed and filtered values of the local level
fit_kf <- fkf(a0 = y[1], 
              P0 = matrix(100),
              dt = matrix(0),
              ct = matrix(0),
              Tt = matrix(1),
              Zt = matrix(1),
              HHt = matrix(1385.066),
              GGt = matrix(15124.13),
              yt = t(y))

fit_ks <- fks(fit_kf)

# Filtered estimates
mu_t <- fit_kf$att[1,]
mu_t <- ts(mu_t)
tsp(mu_t) <- tsp(y)

# Smoothed estimates
mu_T <- fit_ks$ahatt[1,]
mu_T <- ts(mu_T)
tsp(mu_T) <- tsp(y)

```



```{r eval=T, echo=F}
plot.ts(
  y,
  col = "darkgrey",
  xlab = "",
  ylab = "",
  lwd = 1.5,
  main = "Flow of the River Nile and its Estimated Smoothed and Filtered Levels"
)
lines(mu_t, col = "black")
lines(mu_T, col = "blue")
legend(
  "bottomleft",
  legend = c("Flow", "Filtered Level", "Smoothed Level"),
  lwd = c(2, 1),
  col = c("darkgrey", "black", "blue"),
  bty = "n"
)

```


To illustrate the weight decomposition, consider the estimate of the local level at time $t = 50$. Koopman and Harvey (2003) showed that the smoothed estimate $\alpha_{t \mid T}$ can be expressed as

$$
\alpha_{t \mid T} = \sum_{j=1}^{T} w_j(\alpha_{t \mid T}) y_j,
$$

where $w_j(\alpha_{t \mid T})$ denotes the observation weight assigned to $y_j$. Missing values (`NA`) may be present in the data; in such cases, the corresponding weights are set to zero.

Similarly, the filtered estimate can be written as


$$
\alpha _{t|t}=\sum_{j=1}^{t}w_{j}(\alpha _{t|t})y_{j}.
$$



We can compute the weight assigned to each observation using the `wex` function and compare the resulting weighted averages with the corresponding estimates obtained from the Kalman filter and smoother.

```{r}
wts <- wex(Tt=matrix(1),
        Zt=matrix(1),
        HHt = matrix(1385.066),
        GGt = matrix(15124.13),
        yt = t(y),
        t=50)
```


We can also visualize the weights assigned to each observation.

```{r}
par(mfrow = c(2, 1),
    mar = c(2.2, 2.2, 1, 1),
    cex = 0.8)
plot(
  wts$Wt,
  col = "darkgrey",
  xlab = "",
  ylab = "",
  lwd = 1.5,
  type="l",
  main="Filtering weights"
)
plot(
  wts$WtT,
  col = "blue",
  xlab = "",
  ylab = " ",
  lwd = 1.5,
  type="l",
  main="Smoothing weights"
)

```

Finally, we verify that the filtered and smoothed estimates obtained from the Kalman filter coincide with those computed using the observation weights.

```{r}
cat(
  "\nSmoothed level computed using the weights = ",
  sum(y * as.numeric(wts$WtT), na.rm = TRUE),
  "\nSmoothed level from the Kalman smoother = ",
  fit_ks$ahatt[50]
)
```

```{r}
cat(
  "\nFiltered level computed using the weights = ",
  sum(y * as.numeric(wts$Wt), na.rm = TRUE),
  "\nFiltered level from the Kalman filter = ",
  fit_kf$att[50]
)
```

## Example 2: Decomposing Variable Contributions in a Dynamic Factor Model

In this example, we show how to compute observation weights in a dynamic factor model (DFM) and use them to decompose the latent factor into contributions from individual variables.

More formally, let $x_t = (x_{1,t}, x_{2,t}, \dots, x_{n,t})^{\prime}$, for $t = 1, 2, \dots, T$, denote a vector of $n$ monthly series that have been transformed to achieve stationarity and standardized. A dynamic factor model assumes that $x_t$ can be decomposed into two unobserved orthogonal components representing common and idiosyncratic factors. The model is given by

$$
x_t = \Lambda F_t + \varepsilon_t, \hspace{2pt} \varepsilon_t \sim N(0, R),
$$

where $F_t$ is an $(r \times 1)$ vector of unobserved common factors, $\Lambda$ is an $(n \times r)$ matrix of factor loadings, and $\varepsilon_t$ is an $(n \times 1)$ vector of idiosyncratic components. The common factors are assumed to follow the stationary VAR($p$) process

$$
F_t = \sum_{s=1}^{p} \Phi_s F_{t-s} + u_t, \hspace{2pt} u_t \sim N(0, Q),
$$

where $\Phi_s$ are $(r \times r)$ matrices of autoregressive coefficients. Estimation and signal extraction can then be carried out using standard Kalman filtering and smoothing methods.

In this illustrative example, we use a dataset containing 10 monthly economic indicators spanning January 2000 to November 2021. All variables have been log-differenced, when necessary, to achieve stationarity. We assume a single latent factor following an AR(1) process.

The standardized data series and the corresponding smoothed factor estimate are shown in the plot below.

```{r echo = T}
# Use the indicators dataset from wex
df <- scale(wex::indicators[,-1])

# Define the state-space matrices
Zt <- matrix(c(0.37873307, 0.37438154, 0.37767322,
                 0.02433999, 0.36020426, 0.23031769,
                 0.36584474, 0.35066644, 0.33420247,
                 0.01379571),
          ncol=1
)

Tt <- matrix(-0.3676422)

HHt <- matrix(1)

GGt <- matrix(c(
  0.7891011, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 0.3235747, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 0, 0.7673983, 0, 0, 0, 0, 0, 0, 0,
  0, 0, 0, 0.01704776, 0, 0, 0, 0, 0, 0,
  0, 0, 0, 0, 0.9979156, 0, 0, 0, 0, 0,
  0, 0, 0, 0, 0, 0.8496217, 0, 0, 0, 0,
  0, 0, 0, 0, 0, 0, 0.8132641, 0, 0, 0,
  0, 0, 0, 0, 0, 0, 0, 0.9084006, 0, 0,
  0, 0, 0, 0, 0, 0, 0, 0, 0.7601053, 0,
  0, 0, 0, 0, 0, 0, 0, 0, 0, 0.1789897
), nrow = 10, ncol = 10, byrow = TRUE)

# Computing the smoothed values of the latent factor

fit_kf <- fkf(a0 = 0, 
              P0 = diag(1e6, 1),
              dt = matrix(0,1),
              ct = rep(0,ncol(df)),
              Tt = Tt,
              Zt = Zt,
              HHt = HHt,
              GGt = GGt,
              yt = t(df))

fit_ks <- fks(fit_kf)


df.ts <- ts(df,start=c(2000,1),frequency = 12)
f_t <- ts(fit_ks$ahatt[1,])
tsp(f_t) <- tsp(df.ts)


plot(df.ts,plot.type = "single",col="darkgrey",xlab="Year",ylab="")
abline(h=0,lty=1)
lines(f_t, col="blue")
legend(
  "bottomleft",
  legend = c("Standardized series", "Smoothed factor estimate"),
  lwd = c(1, 1),
  col = c("darkgrey", "blue"),
  bty = "n"
)
```

We now use the `wex` function to decompose the final estimate of the latent factor into contributions from each observed variable.


```{r}
# Extract weights for the final observation
wts <- wex(Tt=Tt,
        Zt=Zt,
        HHt = HHt,
        GGt = GGt,
        yt = t(df),
        t=nrow(df))
# Compute contributions

# Extract smoothing weights corresponding to the target state
sweights <- t(wts$WtT[1, , ])
colnames(sweights) <- colnames(df)

# Compute variable contributions as weighted sums of the observed data
contributions <- colSums(sweights * df, na.rm = TRUE)

```

The contributions are summarized in the Table below:

```{r echo=F}
names <- c("Total industrial production in Israel",
        "Trade revenue",
        "Service revenue",
        "Employment (excluding absent workers)",
        "Exports of services",
        "Building starts",
        "Imports of consumer goods",
        "Imports of production inputs",
        "Exports of goods",
        "Job openings")

df.contributions <- data.frame(Variable=names,
                               Contribution=as.numeric(contributions))
df.contributions <-rbind(df.contributions,
      data.frame(Variable="Total",Contribution=sum(df.contributions$Contribution)))

df.contributions$Contribution <- round(df.contributions$Contribution,3)

df.contributions %>%
  kbl(caption = "Varibale Contributions") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```


## Implementation

The package supports two computational backends:

- **FKF**: fast and efficient, but requires non-singular prediction error covariance matrices.
- **KFAS**: more flexible, allowing for singular prediction error covariance matrices.


# References

- Koopman, S. J., and Harvey, A. C. (2003). Computing observation weights for signal extraction and filtering. *Journal of Economic Dynamics and Control*, 27(7), 1317–1333. https://doi.org/10.1016/S0165-1889(02)00061-1

- Helske, J. (2017). KFAS: Exponential Family State Space Models in R. *Journal of Statistical Software*, 78(10), 1–39. https://doi.org/10.18637/jss.v078.i10



# Disclaimer

The views expressed here are solely of the author and do not necessarily represent the views of the Bank of Israel. 

Please note that `wex` is still under development and may contain bugs or other issues that have not yet been resolved. While we have made every effort to ensure that the package is functional and reliable, we cannot guarantee its performance in all situations. 

We strongly advise that you regularly check for updates and install any new versions that become available, as these may contain important bug fixes and other improvements. By using this package, you acknowledge and accept that it is provided on an "as is" basis, and that we make no warranties or representations regarding its suitability for your specific needs or purposes.

