---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wex

<!-- badges: start -->
<!-- badges: end -->

`wex` is an R package designed to compute the exact observation weights for the Kalman filter and smoother using the method described in Koopman and Harvey (2003).
It offers an in-depth exploration of state-space models, enabling researchers and practitioners to extract valuable insights from their time series data. This functionality is particularly useful in dynamic factor models, where the weights can be employed to decompose the contributions of individual variables to the latent factors (see Example 2 below).



## Installation

You can install the development version of wex from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("timginker/wex")
```

## Example 1: Local level model

In this illustrative example, we fit the local level model to the Nile dataset and compute the associated smoothed and filtered values. 

The resulting estimates are presented in the plot below:

```{r echo=F, eval=T}
library(FKF)
library(wex)
library(kableExtra)
data(Nile)
y <- Nile
y[c(3, 10)] <- NA # NA values can be handled

# Computing smoothed and filtered values of the local level
fit_kf <- fkf(a0 = y[1], 
              P0 = matrix(100),
              dt = matrix(0),
              ct = matrix(0),
              Tt = matrix(1),
              Zt = matrix(1),
              HHt = matrix(1385.066),
              GGt = matrix(15124.13),
              yt = t(y))

fit_ks <- fks(fit_kf)

# Filtered values of the local level
mu_t <- fit_kf$att[1,]
mu_t <- ts(mu_t)
tsp(mu_t) <- tsp(y)

# Smoothed values of the local level
mu_T <- fit_ks$ahatt[1,]
mu_T <- ts(mu_T)
tsp(mu_T) <- tsp(y)

```



```{r eval=T, echo=F}
plot.ts(
  y,
  col = "darkgrey",
  xlab = "",
  ylab = "",
  lwd = 1.5,
  main = "Flow of the River Nile and its Estimated Smoothed and Filtered Levels"
)
lines(mu_t, col = "black")
lines(mu_T, col = "blue")
legend(
  "bottomleft",
  legend = c("Flow", "Filtered Level", "Smoothed Level"),
  lwd = c(2, 1),
  col = c("darkgrey", "black", "blue"),
  bty = "n"
)

```
Now, w.l.o.g., let's consider the 50th value of the estimated local level

```{r echo=F, eval=T}
cat("smoothed level[50] = ",mu_T[50])
```
Koopman and Harvey (2003) showed that the smoothed estimates can be represented as:

```math
$$
\alpha _{t|T}=\sum_{j=1}^{T}w_{j}(\alpha _{t|T})y_{j}.
$$
```

Similarly, the filtered estimated can be computed as:

```math
$$
\alpha _{t|t}=\sum_{j=1}^{t}w_{j}(\alpha _{t|Tt})y_{j}.
$$
```


Now, We can compute the weight of each observation using the `wex` function, and compare the local level estimates obtained from the weighted average of the observed data with the associated estimates obtained from the Kalman filter and smoother.

```{r}
wts=wex(Tt=matrix(1),
        Zt=matrix(1),
        HHt = matrix(1385.066),
        GGt = matrix(15124.13),
        yt = t(y),
        t=50)
```


We can also visualize the weights assigned to each observation:

```{r}
par(mfrow = c(2, 1),
    mar = c(2.2, 2.2, 1, 1),
    cex = 0.8)
plot(
  wts$Wt,
  col = "darkgrey",
  xlab = "",
  ylab = "",
  lwd = 1.5,
  type="l",
  main="Filtering weights"
)
plot(
  wts$WtT,
  col = "blue",
  xlab = "",
  ylab = " ",
  lwd = 1.5,
  type="l",
  main="Smoothing weights"
)

```

It is also easy to verify the identity between the smoothed and filtered levels obtained from the Kalman filter and the corresponding estimates computed using the weights.

```{r}
cat("\n Smoothed level computed using the weights = ",
    sum(y*as.numeric(wts$WtT),na.rm = T),
    " \n Smoothed level from the Kalman Filter = ",fit_ks$ahatt[50])
```

```{r}
cat("\n Filtered level computed using the weights = ",
    sum(y*as.numeric(wts$Wt),na.rm = T),
    " \n Filtered level from the Kalman Filter = ",fit_kf$att[50])
```

## Example 2:

In this example we show how to compute observation weights for a Dynamic Factor Model (DFM) which are then used to decompose the latent factor into the contribution of each variable.

More formally, let $x_{t}=(x_{1,t},x_{2,t},...,x_{n,t})^{\prime}$ with $t =1,2,...,T$ be a vector of $n$ monthly series that have been transformed to become stationary and standardized. A dynamic factor model (DFM) assumes that it is possible to decompose $x_{t}$ into two unobserved orthogonal components representing common and idiosyncratic factors. The model is specified as follows:

```math
$$
x_{t}=\Lambda F_{t}+\varepsilon_{t}, \hspace{2pt} \varepsilon_{t}\sim N(0,R),
$$
```
where $F_{t}$ is an $(r \times 1)$ vector of unobserved common factors, $\Lambda$ is an $(n \times r)$ matrix of their loadings, and $\varepsilon_{t}$ is an $(n \times 1)$ vector of the idiosyncratic components. The factors are assumed to have the following stationary VAR(p) representation:


```math
$$
F_{t}=\sum_{s=1}^{p}\Phi _{s}F_{t-s}+u_{t},\hspace{2pt} u_{t}\sim N(0,Q)
$$
```
where $\Phi_{s}$ are $(r \times r)$ matrices of autoregressive coefficients. The related inference and forecast procedures can be carried out using the standard Kalman filter techniques.

In this illustrative example, we use a dataset containing 10 monthly economic indicators covering the period from January 2000 to November 2021. All variables have been log-differenced, when necessary, to achieve stationarity. We assume that there is a single laten factor that follows an AR(1) process.

The normalized data series are summarized in the plot below:

```{r echo = F}
# using the indicators dataset from wex
df <- scale(wex::indicators[,-1])
df.ts <- ts(df,start=c(2000,1),frequency = 12)

plot(df.ts,plot.type = "single",col="darkgrey",xlab="Year",ylab="")
abline(h=0,lty=1)
```

Below, we use the `wex` function to decompose the last value of the latent factor into a contribution of each variable.


```{r}
# defining the stat-space matrices
Zt <- matrix(c(0.37873307, 0.37438154, 0.37767322,
                 0.02433999, 0.36020426, 0.23031769,
                 0.36584474, 0.35066644, 0.33420247,
                 0.01379571),
          ncol=1
)

Tt <- matrix(-0.3676422)

HHt <- matrix(1)

GGt <- matrix(c(
  0.7891011, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 0.3235747, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 0, 0.7673983, 0, 0, 0, 0, 0, 0, 0,
  0, 0, 0, 0.01704776, 0, 0, 0, 0, 0, 0,
  0, 0, 0, 0, 0.9979156, 0, 0, 0, 0, 0,
  0, 0, 0, 0, 0, 0.8496217, 0, 0, 0, 0,
  0, 0, 0, 0, 0, 0, 0.8132641, 0, 0, 0,
  0, 0, 0, 0, 0, 0, 0, 0.9084006, 0, 0,
  0, 0, 0, 0, 0, 0, 0, 0, 0.7601053, 0,
  0, 0, 0, 0, 0, 0, 0, 0, 0, 0.1789897
), nrow = 10, ncol = 10, byrow = TRUE)

# Extracting weights for the last obseravtion
wts <- wex(Tt=Tt,
        Zt=Zt,
        HHt = HHt,
        GGt = GGt,
        yt = t(df),
        t=nrow(df))
# Computing contributions

ctr <- matrix(0,nrow=nrow(df),ncol=ncol(df))

for (i in 1:nrow(df)) {
  
  ctr[i,]=as.numeric(wts$WtT[,,i]*df[i,])
  
}

ctr <- data.frame(ctr)
colnames(ctr) <-colnames(df)

contributions <- colSums(ctr,na.rm=T)
```

The contributons are summarized in the Table below:

```{r echo=F}
names <- c("Total industrial production in Israel",
        "Trade revenue",
        "Service revenue",
        "Employment (excluding absent workers)",
        "Exports of services",
        "Building starts",
        "Imports of consumer goods",
        "Imports of production inputs",
        "Exports of goods",
        "Job openings")

df.contributions <- data.frame(Variable=names,
                               Contribution=as.numeric(contributions))
df.contributions <-rbind(df.contributions,
      data.frame(Variable="Total",Contribution=sum(df.contributions$Contribution)))

df.contributions$Contribution <- round(df.contributions$Contribution,3)

df.contributions %>%
  kbl(caption = "Varibale Contributions") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```


# References

Koopman, S. J., & Harvey, A. (2003). Computing observation weights for signal extraction and filtering. Journal of Economic Dynamics and Control, 27(7), 1317-1333.


# Disclaimer

The views expressed here are solely of the author and do not necessarily represent the views of the Bank of Israel. 

Please note that `wex` is still under development and may contain bugs or other issues that have not yet been resolved. While we have made every effort to ensure that the package is functional and reliable, we cannot guarantee its performance in all situations. 

We strongly advise that you regularly check for updates and install any new versions that become available, as these may contain important bug fixes and other improvements. By using this package, you acknowledge and accept that it is provided on an "as is" basis, and that we make no warranties or representations regarding its suitability for your specific needs or purposes.

